An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-31 scagnostics.Rmd
Summary table symbols updated.
Notation updated.

The term "projection pursuit" was coined by @ff74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($k=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $k$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $k$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis.

@@

Our motivation is based on applications in physics, to aid the interpretation of model fits on experimental results. Consider a physical model with a set of $m$ free parameters, that cannot be measured directly and are determined by fitting a set of $p$ experimental observations, for which predictions can be calculated when fixing all $m$ parameters. Note here, that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numerical computation. Moreover, a single prediction can be a complicated function of all free parameters. In addition, we often deal with large number of observations, $p \sim 100-1000$ as well as sizable parameter spaces $m \sim 10$. The results of a fit are generally interpreted using combinations of variables selected by intuition and prior knowledge. But this begs the question, whether important associations are missed because tools to search are not available.

@@

- Definition of pp index
- Optimisation
- Illustration of what is already available in tourr
- New index explanations
-->
A projection pursuit index (PPI) is a scalar function $f$ defined on an $n$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate more interesting distribution, and therefore maximising $f$ over all possible $n$ dimensional projections of a data set with $p>n$ parameters will find the most interesting projections. As most indices characterise deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterising differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterise it by measures based on e.g. the minimal spanning tree or the mutual information.

@@

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $n$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuatins or a consequence of noisy index functions, and @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimisation for the identification of the maximum. On the other hand the global maximum may not be the only projection of interest, and one may want to observe the selected views in the context of the full distribution rather than a static view. These issues are addressed when combining projection pursuit with the grand tour [@CBCH94]. In this case the properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshoothing and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible alogrithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. In addition the tourr package provides the \texttt{search\_geodesic} search function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

## New index functions
\label{sec:indexDef}
Focusing now on $n=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developped for the selection of variable pairs that show interesting features in a scatter plot, but we generalise the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimising over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:\footnote{We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.}

@@

- **pipe**: $p$ points where $n-1$ points are independently drawn from a uniform distribution between $[-1,1]$, and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- **sine**:  $p$ points where $n-1$ points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"
- **sphere**:  $p$ points on an n-dimensional sphere randomly selected using the geozoo package [@geozoo], as an example of dataset without special views, labelled "sphere"

Two samples sizes are also used: $p=100$ (small) and $p=1000$ (large).

@@

- index values for pair of variables, explicitly comparing the uninformative and special view of each dataset, as well as discussing the dependence on the sample size, see Section \ref{sec:dataOv}
- tour between pairs of variables, considering a planned tour that moves from the $V_1 - V_2$ view either to the $V_3 - V_4$ view (no special structure encountered over the full path) or to the $V_5 - V_6$ (moving from uninformed to special view), see Section \ref{sec:smooth}
- optimisation path, starting from $V_1 - V_2$, how can we use the available index functions and optimisation algorithms to identify the special view, see Section \ref{sec:guided}
- rotating a dataset with functional dependence to evaluate the rotation dependence of the index functions, see Section \ref{sec:rot}
- benchmarking the computing time in Section \ref{sec:timer}

@@

To test the behaviour of the index functions we use three types of simulated data sets based on different underlying distributions, and with different built-in features. They are defined in $n=6$ dimensions with $p$ randomly selected points $x_i, i = 1,..,n$ according to:

- sphere:  $p$ points on an n-dimensional sphere randomly selected using the geozoo package [@geozoo], as an example of dataset without special views, labelled "sphere"
- pipe: $p$ points where $n-1$ points are independently drawn from a uniform distribution between $[-1,1]$, and enforcing $x_n^2 + x_{n-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- sine:  $p$ points where $n-1$ points are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_n = \sin(x_{n-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"

For each distribution we consider a small sample of $p=100$ and a large sample with $p=1000$.

@@

To get an overview of the dataset we show views of the first two parameters showing the independent distribution and the last two parameters showing the special view. The enforced relationships will also affect the 1-d densities as can be seen from \ref{fig:dataPlotsDensity}, which means that views including sizable fractions of V6 (and V5 for the Pipe distribution) will be distinct from uninformed views of distributions 2 and 3.

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the first two parameters V1 and V2, i.e. showing a typical uninformed view for each dataset."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the last two parameters V5 and V6, i.e. showing the special view built into the Pipe and Sine datasets."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

@@

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters V5 and V6 are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine dataset the parameter V6 follows a very distinct distribution. Note also that the first five paramters of the Sine dataset appear to be drawn from different normal distributions, which is a result of the rescaling, shifting the modes to different values for each direction."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- melt(ds)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

We next compare the various index values for three views: V1 vs V2, V5 vs V6 and V1 vs V6 (capturing differences from the different distribution in V6). For the Sphere distribution by construction we see only random fluctuations between the different views, which can be sizable in particular for small datasets. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Moreover, minimising the convex index (or maximising 1-convex) should also allow us to detect the special view. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fictional dependence.

@@

### Random fluctuations
We first study the smoothness of each measure calculated on a sequence of 2-d projections obtained via an interpolated planned tour path moving between the projection onto V1-V2 to the projection onto V3-V4 and using the default interpolation angle of 0.05. Each time step refers to an interpolation step. Since the value of the TIC index is strongly dependent on the input distribution sample size we have normalised the TIC measure by its maximum for each individual dataset, all other measures are reproduced as reported by the calculation. The results are shown in Figure \ref{fig:plotV1V2toV3V4}. The standard index functions "holes" and "cmass" show the desired behaviour in that the change is smooth over interpolated views. Considering now the indices from the scagnostics family, we first note stronge dependence on the sample size. The convex index is largely sensitive to random fluctuations when considering small samples (p=100), while being mostly flat for larger samples, especially in the case of an underlying uniform distribution. The skinny index on the other hand is systematically larger for the smaller samples, while stringy appears not to be sensitive to the sample size, but resulting in large fluctuations and sharp jumps for all datasets. Distance correlation and the spline based index show similar behaviour, i.e. when considering large enough data samples (p=1000) the values are consistently close to zero, as expected for a distribution without structure. On the other hand somewhat larger values (but still far from the maximum value of 1) are found when considering small datasets, where distributions are however mostly smooth, with few kinks depending on index function and underlying distribution. Finally the information based metrics also show systematic differences based on the sample size, with sampling fluctuations being significant for small datasets. Note here that fluctuations present in Figure \ref{fig:plotV1V2toV3V4} may be amplified, as maximum values for most index functions are far from the theoretic maximum.

@@

### Moving into informative projections
We repeat the same exercise, but now we are interested in how the index values change when going from the view in V1 vs V2 to the view in V5 vs V6, which will also allow us to view the evolution between typical low and high values of the index function. Results are shown in Figure \ref{fig:plannedtour}.
We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparent much closer to the final projection, making this structure more challenging to detect with a guided tour.
Various indices appear less noisy than seen in \ref{fig:plotV1V2toV3V4}, but note that the y-axis scale is different (typically reaching higher index values in this example, whereas smaller fluctuations are more amplified in the previous views).

@@

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="As Figure \\ref{fig:plotV1V2toV3V4} but moving from the projection onto V1-V2 to the projection onto V5-V6, i.e. starting from random distribution view, and moving into the special view."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

## Distributions of index values
\label{sec:dist}
Compare typical index values over the two tour paths. Notation: path 1 / 2 labels tour from V1-V2 to V3-V4 / V5-V6. Typically find broader distributions for small samples because of the large fluctuations in this case. Some indices show preferred behaviour, i.e. sharp peaks around low values for path 1, and very broad distributions leading up to maximum values for path 2. For other indices we can see the strong dependence of the index on sample size, the most extreme example is the skinny index.

@@

## Rotation dependence
\label{sec:rot}
\textcolor{red}{testing rotation dependence by turning sine dataset with $\alpha \in [0,2\pi]$}
Interesting to note that some indexes are symmetric around $\pi$, but scagnostics indexes are very arbitrary, dependence is introduced via binning (and outlier removal?) here. Observe huge drops for some of the indexes! 
```{r rotationDep, fig.width=6, fig.height=7, fig.cap="Rotation dependence of the considered index functions tested on the sine 1000 dataset, by rotating the special view by an angle alpha."}
if(!file.exists("cache/rotationDep.rda")){
  sineM <- as.matrix(select(sin1000, V5, V6))
  sc <- tibble(
      holes=numeric(),
      cmass=numeric(),
      convex=numeric(),
      skinny=numeric(),
      stringy=numeric(),
      dcor2d=numeric(),
      splines2d=numeric(),
      MIC=numeric(),
      TIC=numeric(),
      alpha=numeric()
  )
  for (a in seq(0,2*pi, pi/100)){
    rotM <- matrix(c(cos(a), sin(a), -sin(a), cos(a)), ncol = 2)
    dprj <- sineM %*% rotM
    scagRes <- scagnostics(dprj)
    dcorRes <- dcor2d(dprj[,1], dprj[,2])
    splineRes <- splines2d(dprj[,1], dprj[,2])
    mineRes <- mine(dprj[,1], dprj[,2])
    holesRes <- holes(dprj)
    cmassRes <- cmass(dprj)
    sc <- add_row(sc, holes=holesRes, cmass=cmassRes,
                  convex=scagRes[,"Convex"], skinny=scagRes[,"Skinny"],
                  stringy=scagRes[,"Stringy"], dcor2d=dcorRes,
                  splines2d=splineRes, MIC=mineRes$MIC,
                  TIC=mineRes$TIC, alpha=a)
  }
  save(sc, file = "cache/rotationDep.rda")
} else {
  load("cache/rotationDep.rda")
}

@@

### Scagnostics binning
The implementation of scagnostics will always perform a binning step first, before evaluating the measures. The binning will introduce random fluctuations on interpolated tour paths, but we may also consider using it to reduce fluctuations from the discreteness inherent to the index definitions, see in particular the definition of the stringy index. This needs some care however as different values may be preferred for different structures and depending on which index is being considered. Moreover, choosing a binning that is too rough may result in loss of discrimination power as structures can no longer be resolved. As an example we compare in Figure \ref{fig:sprialScagBinning} the scagnostics index traces for the Spiral 1000 dataset when moving from V1-V2 to V5-V6, for three values of the bins parameter entering the hexagonal binning procedure. We find indeed that different values may be preferred depending on the index considered, with the default bins=50 being the better choice for the convex and skinny index, while bins=20 may work better for the stringy index. Note that neither choice results in smooth curves for any of the indexes considered here.

@@

\begin{table}
\begin{center}
\caption{Summary of findings, showing to what extend the considered index functions pass the criteria for a good PPI. Checkmark symbols good behaviour, exclamation mark symbols some issues while cross symbols failure on the corresponding criteria. Each index has particular strengths and drawbacks and selection must be guided by the considered example, see text for details.}
\label{tab:summary}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Index & smooth & squintability & flexible & rotation invariant & speed \\
\hline
Convex & ! & ! & ! & ! & \checkmark \\
Skinny & ! & !& ! & ! & \checkmark \\
Stringy & $\times$ & $\times$ & ! & ! & \checkmark \\
\hline
splines2D & \checkmark & \checkmark & ! & $\times$ & \checkmark \\
dcor2D & \checkmark & \checkmark & ! & $\times$ & ! \\
\hline
MINE & ! & \checkmark  & \checkmark & ! & ! \\
\hline
\end{tabular}
\end{center}
\end{table}

>>>>

The term "projection pursuit" was coined by @ff74 to describe a procedure for searching high (say $p-$)dimensional data for "interesting" low-dimensional projections ($d=1$ or $2$ usually). The procedure, originally suggested by @kr69, involves defining a criterion function, or index, that measures the "interestingness" of each $d$-dimensional projection of $p$-dimensional data. This criterion function is optimized over the space of all $d$-dimensional projections of $p$-space, searching for both global and local maxima. It is hoped that the resulting solutions reveal low-dimensional structure in the data not found by methods such as principal component analysis.

@@

Our motivation is based on applications in physics, to aid the interpretation of model fits on experimental results. Consider a physical model with a set of $p$ free parameters, that cannot be measured directly and are determined by fitting a set of $n$ experimental observations, for which predictions can be calculated when fixing all $p$ parameters. Note here, that while we often have analytic expressions for the predictions, this is not always the case and we may have to rely on numerical computation. Moreover, a single prediction can be a complicated function of all free parameters. In addition, we often deal with large number of observations, $n \sim 100-1000$ as well as sizable parameter spaces $p \sim 10$. The results of a fit are generally interpreted using combinations of variables selected by intuition and prior knowledge. But this begs the question, whether important associations are missed because tools to search are not available.

@@

- Definition of pp index
- Optimisation
- Illustration of what is already available in tourr
- New index explanations
-->
A projection pursuit index (PPI) is a scalar function $f$ defined on an $d$-dimensional data distribution. Typically the definition is such that larger values of $f$ indicate more interesting distribution, and therefore maximising $f$ over all possible $d$ dimensional projections of a data set with $p>d$ parameters will find the most interesting projections. As most indices characterise deviations from a normal distribution, one would first map the empiric data distribution onto a density for which such deviations can then be defined in different ways, see e.g. @CBC93. When departing from the idea of characterising differences to the normal distribution it may however be preferred to work directly with the empirical data distribution and instead characterise it by measures based on e.g. the minimal spanning tree or the mutual information.

@@

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuatins or a consequence of noisy index functions, and @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimisation for the identification of the maximum. On the other hand the global maximum may not be the only projection of interest, and one may want to observe the selected views in the context of the full distribution rather than a static view. These issues are addressed when combining projection pursuit with the grand tour [@CBCH94]. In this case the properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshoothing and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible alogrithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. In addition the tourr package provides the \texttt{search\_geodesic} search function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

## New index functions
\label{sec:indexDef}
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developped for the selection of variable pairs that show interesting features in a scatter plot, but we generalise the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimising over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:\footnote{We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.}

@@

- **pipe**: $n$ points where all $p$ directions are independently drawn from a uniform distribution between $[-1,1]$, but enforcing $x_p^2 + x_{p-1}^2 = 1 \pm 0.1$ by rejection sampling as an example of a dataset with special view that is not described by functional dependence, and where sampling boundaries are apparent in the distribution, labelled "pipe"
- **sine**:  $n$ points where $p-1$ directions are independently drawn from a normal distribution with mean 0 and variance 1, and with $x_p = \sin(x_{p-1}) + \mathrm{jittering}$ as an example dataset with special view described by a function, labelled "sine"
- **spiral**:  $n$ points where $p-2$ directions are independently drawn from a normal distribution, and direction $p-1$ and $p$ are sampled on an Archimedean spiral, i.e. $r = a + b \theta$, with $a=b=0.1$ and we sample angles $\theta$ from a normal distribution with mean 0 and variance $2\pi$, giving a spiral with higher densities at lower radii. We take the absolute value of $\theta$ to fix the direction of the spiral shape when sampling the points.

Two samples sizes are also used: $n=100$ (small) and $n=1000$ (large).

@@

- index values for pair of variables, explicitly comparing the uninformative and special view of each dataset, as well as discussing the dependence on the sample size, see Section \ref{sec:dataOv}
- tour between pairs of variables, considering a planned tour that moves from the $x_1 - x_2$ view either to the $x_3 - x_4$ view (no special structure encountered over the full path) or to the $x_5 - x_6$ (moving from uninformed to special view), see Section \ref{sec:smooth}
- optimisation path, starting from $x_1 - x_2$, how can we use the available index functions and optimisation algorithms to identify the special view, see Section \ref{sec:guided}
- rotating a dataset with functional dependence to evaluate the rotation dependence of the index functions, see Section \ref{sec:rot}
- benchmarking the computing time in Section \ref{sec:timer}

@@

@@

To get an overview of the dataset we show views of the first two directions showing the independent distribution and the last two directions showing the special view. The enforced relationships will also affect the 1-d densities as can be seen from \ref{fig:dataPlotsDensity}, which means that views including sizable fractions of $x_6$ (and $x_5$ for the Pipe and Spiral distribution) will be distinct from uninformed views of distributions 2 and 3.

```{r dataPlotsV1V2, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the first two parameters $x_1$ and $x_2$, i.e. showing a typical uninformed view for each dataset."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V1, V2)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

```{r dataPlotsV5V6, fig.width=6, fig.height=7, fig.cap="Projection of all data sets considered onto the last two parameters $x_5$ and $x_6$, i.e. showing the special view built into the datasets."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  pC <- ggplot(ds, aes(V5, V6)) +
    geom_point() +
    ggtitle(pLabels[i]) + theme(aspect.ratio=1)
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

@@

```{r dataPlotsDensity, fig.width=6, fig.height=7, fig.cap="One dimensional density distribution for all parameters and all data sets considered. Apart from statistical fluctuations we see that for the Pipe datasets the parameters $x_5$ and $x_6$ are pushed towards more extreme values by vetoing points away from the circle outline, and for the Sine dataset the parameter $x_6$ follows a very distinct distribution. Note also that the first five paramters of the Sine dataset appear to be drawn from different normal distributions, which is a result of the rescaling, shifting the modes to different values for each direction."}
pL <- list()
i <- 1
pLabels <- c("Spiral 100", "Spiral 1000", "Pipe 100", "Pipe 1000", "Sine 100", "Sine 1000")
for(ds in list(spiral100, spiral1000, pipe100, pipe1000, sin100, sin1000)){
  distData <- melt(ds)
  pC <- ggplot(distData, aes(value, color=variable)) +
    geom_density() +
    ggtitle(pLabels[i])
  pL[[i]] <- pC
  i <- i+1
}
grid.arrange(pL[[1]], pL[[2]], pL[[3]], pL[[4]], pL[[5]], pL[[6]],
             ncol=2)
```

We next compare the various index values for three views: $x_1$ vs $x_2$, $x_5$ vs $x_6$ and $x_1$ vs $x_6$ (capturing differences from the different distribution in $x_6$). For the Sphere distribution by construction we see only random fluctuations between the different views, which can be sizable in particular for small datasets. Considering next the Pipe distribution, we see that skinny, stringy and dcor2d are increased for the special view, but in particular MIC and TIC index are much larger and may be used to detect this type of structure in the dataset. Moreover, minimising the convex index (or maximising 1-convex) should also allow us to detect the special view. Finally the Sine distribution shows that several of the indices are maximised in the special view, in particular also the splines2d index designed for the detection of fictional dependence.

@@

### Random fluctuations
We first study the smoothness of each measure calculated on a sequence of 2-d projections obtained via an interpolated planned tour path moving between the projection onto $x_1$-$x_2$ to the projection onto $x_3$-$x_4$ and using the default interpolation angle of 0.05. Each time step refers to an interpolation step. Since the value of the TIC index is strongly dependent on the input distribution sample size we have normalised the TIC measure by its maximum for each individual dataset, all other measures are reproduced as reported by the calculation. The results are shown in Figure \ref{fig:plotV1V2toV3V4}. The standard index functions "holes" and "cmass" show the desired behaviour in that the change is smooth over interpolated views. Considering now the indices from the scagnostics family, we first note stronge dependence on the sample size. The convex index is largely sensitive to random fluctuations when considering small samples ($n=100$), while being mostly flat for larger samples, especially in the case of an underlying uniform distribution. The skinny index on the other hand is systematically larger for the smaller samples, while stringy appears not to be sensitive to the sample size, but resulting in large fluctuations and sharp jumps for all datasets. Distance correlation and the spline based index show similar behaviour, i.e. when considering large enough data samples ($n=1000$) the values are consistently close to zero, as expected for a distribution without structure. On the other hand somewhat larger values (but still far from the maximum value of 1) are found when considering small datasets, where distributions are however mostly smooth, with few kinks depending on index function and underlying distribution. Finally the information based metrics also show systematic differences based on the sample size, with sampling fluctuations being significant for small datasets. Note here that fluctuations present in Figure \ref{fig:plotV1V2toV3V4} may be amplified, as maximum values for most index functions are far from the theoretic maximum.

@@

### Moving into informative projections
We repeat the same exercise, but now we are interested in how the index values change when going from the view in $x_1$ vs $x_2$ to the view in $x_5$ vs $x_6$, which will also allow us to view the evolution between typical low and high values of the index function. Results are shown in Figure \ref{fig:plannedtour}.
We can see a pretty clear and smooth increase in dcor2d and splines2d as well as MIC and TIC index when approaching the special view of the Sine distribution, which should therefore be efficient to optimise. On the other hand while we also see smooth increase in MIC and TIC when approaching the special view of the Pipe distribution, the increase only becomes apparent much closer to the final projection, making this structure more challenging to detect with a guided tour.
Various indices appear less noisy than seen in \ref{fig:plotV1V2toV3V4}, but note that the y-axis scale is different (typically reaching higher index values in this example, whereas smaller fluctuations are more amplified in the previous views).

@@

```{r plannedtour, fig.width=6, fig.height=7, fig.cap="As Figure \\ref{fig:plotV1V2toV3V4} but moving from the projection onto $x_1$-$x_2$ to the projection onto $x_5$-$x_6$, i.e. starting from random distribution view, and moving into the special view."}
fullResMeltPlanned <- melt(fullResPlanned, id=c("t", "name", "size"))
ggplot(fullResMeltPlanned, aes(x=t, y=value)) +
  geom_line(aes(color=factor(size))) + 
  facet_wrap(variable~name, ncol=3, scales = "free_y", labeller = label_wrap_gen(multi_line=FALSE)) +
  guides(color=FALSE)
```

## Distributions of index values
\label{sec:dist}
Compare typical index values over the two tour paths. Notation: path 1 / 2 labels tour from $x_1$-$x_2$ to $x_3$-$x_4$ / $x_5$-$x_6$. Typically find broader distributions for small samples because of the large fluctuations in this case. Some indices show preferred behaviour, i.e. sharp peaks around low values for path 1, and very broad distributions leading up to maximum values for path 2. For other indices we can see the strong dependence of the index on sample size, the most extreme example is the skinny index.

@@

## Rotation dependence
\label{sec:rot}
\textcolor{red}{testing rotation dependence by turning sine dataset with $\alpha \in [0,2\pi]$, maybe can add second path for spiral?}
Interesting to note that some indexes are symmetric around $\pi$, but scagnostics indexes are very arbitrary, dependence is introduced via binning (and outlier removal?) here. Observe huge drops for some of the indexes! 
```{r rotationDep, fig.width=6, fig.height=7, fig.cap="Rotation dependence of the considered index functions tested on the sine 1000 dataset, by rotating the special view by an angle alpha."}
if(!file.exists("cache/rotationDep.rda")){
  sineM <- as.matrix(select(sin1000, V5, V6))
  sc <- tibble(
      holes=numeric(),
      cmass=numeric(),
      convex=numeric(),
      skinny=numeric(),
      stringy=numeric(),
      dcor2d=numeric(),
      splines2d=numeric(),
      MIC=numeric(),
      TIC=numeric(),
      alpha=numeric()
  )
  for (a in seq(0,2*pi, pi/100)){
    rotM <- matrix(c(cos(a), sin(a), -sin(a), cos(a)), ncol = 2)
    dprj <- sineM %*% rotM
    scagRes <- scagnostics(dprj)
    dcorRes <- dcor2d(dprj[,1], dprj[,2])
    splineRes <- splines2d(dprj[,1], dprj[,2])
    mineRes <- mine(dprj[,1], dprj[,2])
    holesRes <- holes(dprj)
    cmassRes <- cmass(dprj)
    sc <- add_row(sc, holes=holesRes, cmass=cmassRes,
                  convex=scagRes[,"Convex"], skinny=scagRes[,"Skinny"],
                  stringy=scagRes[,"Stringy"], dcor2d=dcorRes,
                  splines2d=splineRes, MIC=mineRes$MIC,
                  TIC=mineRes$TIC, alpha=a)
  }
  save(sc, file = "cache/rotationDep.rda")
} else {
  load("cache/rotationDep.rda")
}

@@

### Scagnostics binning
The implementation of scagnostics will always perform a binning step first, before evaluating the measures. The binning will introduce random fluctuations on interpolated tour paths, but we may also consider using it to reduce fluctuations from the discreteness inherent to the index definitions, see in particular the definition of the stringy index. This needs some care however as different values may be preferred for different structures and depending on which index is being considered. Moreover, choosing a binning that is too rough may result in loss of discrimination power as structures can no longer be resolved. As an example we compare in Figure \ref{fig:sprialScagBinning} the scagnostics index traces for the Spiral 1000 dataset when moving from $x_1$-$x_2$ to $x_5$-$x_6$, for three values of the bins parameter entering the hexagonal binning procedure. We find indeed that different values may be preferred depending on the index considered, with the default bins=50 being the better choice for the convex and skinny index, while bins=20 may work better for the stringy index. Note that neither choice results in smooth curves for any of the indexes considered here.

@@

\begin{table}
\begin{center}
\caption{Summary of findings, showing to what extend the considered index functions pass the criteria for a good PPI. "\checkmark" symbols good behaviour, "$\cdot$" symbols some issues while "$\times$" symbols failure on the corresponding criteria. Each index has particular strengths and drawbacks and selection must be guided by the considered example, see text for details.}
\label{tab:summary}
\begin{tabular}{|l|c|c|c|c|c|c|}
\hline
Index & smooth & squintability & flexible & rotation invariant & speed \\
\hline
Convex & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ & \checkmark \\
Skinny & $\cdot$ & $\cdot$ & $\cdot$ & $\cdot$ & \checkmark \\
Stringy & $\times$ & $\times$ & $\cdot$ & $\cdot$ & \checkmark \\
\hline
splines2D & \checkmark & \checkmark & $\cdot$ & $\times$ & \checkmark \\
dcor2D & \checkmark & \checkmark & $\cdot$ & $\times$ & $\cdot$ \\
\hline
MINE & $\cdot$ & \checkmark  & \checkmark & $\cdot$ & $\cdot$ \\
\hline
\end{tabular}
\end{center}
\end{table}
